---
title: "ICES SAG extraction report"
subtitle: "Time series, reference points, and catch advice by stock and assessment year"
author: "Laurence T. Kell"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
    toc_depth: 2
    df_print: paged
    code_folding: hide
params:
  yrFrom: 2016
  yrTo: 2025
  fetch_scenarios: false
  api_retries: 5
  api_wait_s: 20
  connect_timeout_s: 60
  request_timeout_s: 300
  reuse_stock_list: true
---

```{r setup, include=FALSE}
projectRoot=normalizePath("..", winslash="/", mustWork=TRUE)
knitr::opts_knit$set(root.dir=projectRoot)

knitr::opts_chunk$set(
  echo=FALSE, warning=FALSE, message=FALSE,
  fig.width=10, fig.height=6,
  fig.path="figs/", cache=TRUE, cache.path="cache/",
  include=FALSE)

yrFrom          =params$yrFrom
yrTo            =params$yrTo
assessYrs       =seq(yrTo, yrFrom)
purpose         ="Advice"
statusFilter    ="Published"
fetch_scenarios =isTRUE(params$fetch_scenarios)
apiRetries      =as.integer(params$api_retries)
apiWait         =as.numeric(params$api_wait_s)
connectTimeout  =as.numeric(params$connect_timeout_s)
requestTimeout  =as.numeric(params$request_timeout_s)
reuseStockList  =isTRUE(params$reuse_stock_list)
outDir          =file.path(projectRoot, "data/sag")

dir.create(outDir, recursive=TRUE, showWarnings=FALSE)
dir.create("figs", recursive=TRUE, showWarnings=FALSE)
dir.create("cache", recursive=TRUE, showWarnings=FALSE)

rdataFile=file.path(outDir, sprintf("sag_%d_%d.RData", yrFrom, yrTo))
```

Pulls **ICES Stock Assessment Graph (SAG)** and **Advice View** data for all published advice stocks, retaining separate snapshots for each **assessment year** (`r yrFrom`–`r yrTo`). Extraction runs silently; results appear in the sections below.

| Series / object | Source |
|-----------------|--------|
| SSB, F, recruits, catch | `icesSAG::getSAG(..., data = "summary")` |
| Reference points | `icesSAG::getSAG(..., data = "refpts")` |
| Catch advice | `icesASD::getAdviceViewRecord()` |
| Catch scenarios (optional) | `icesASD::getCatchScenariosTable()` if `fetch_scenarios: true` |

```{r libraries}
suppressPackageStartupMessages({
  library(icesSAG)
  library(icesASD)
  library(plyr)
  library(dplyr)
  library(tidyr)})

sag_messages(FALSE)

if (requireNamespace("httr", quietly=TRUE)) {
  httr::set_config(httr::config(
    connecttimeout=connectTimeout,
    timeout=requestTimeout))
}
```

```{r helpers}
withRetries<-function(expr, label="ICES API", n=apiRetries, wait=apiWait) {
  lastErr=NULL
  for (attempt in seq_len(n)) {
    res=tryCatch(
      expr,
      error=function(e) {
        lastErr<<-e
        NULL})
    if (!is.null(res)) return(res)
    if (attempt < n) {
      warning(label, ": attempt ", attempt, "/", n, " failed (",
              conditionMessage(lastErr), "); retrying in ", wait, " s",
              call.=FALSE)
      Sys.sleep(wait)
      wait=min(wait * 1.5, 120)
    }
  }
  stop(label, " failed after ", n, " attempts: ", conditionMessage(lastErr), call.=FALSE)
}

getCol<-function(df, choices) {
  nms=names(df)
  idx=match(tolower(choices), tolower(nms))
  idx=idx[!is.na(idx)]
  if (length(idx) == 0) NA_character_ else nms[idx[1]]}

c2n<-function(x) suppressWarnings(as.numeric(as.character(x)))

#' Standardise one getSAG() summary table to long SSB / F / catch / recruits.
tidySmry<-function(dat, sid, assYr) {
  if (is.null(dat) || !is.data.frame(dat) || nrow(dat) == 0) {
    return(NULL)}

  yearCol    =getCol(dat, c("year"))
  ssbCol     =getCol(dat, c("ssb", "stocksize", "tbiomass"))
  fCol       =getCol(dat, c("f", "fbar", "fishingpressure"))
  recruitsCol=getCol(dat, c("recruitment", "recruits"))
  catchesCol =getCol(dat, c("catches"))
  landingsCol=getCol(dat, c("landings"))
  discardsCol=getCol(dat, c("discards"))

  if (is.na(yearCol)) return(NULL)

  out=data.frame(
    sid            =sid,
    assYr=assYr,
    year           =c2n(dat[[yearCol]]),
    SSB            =if (!is.na(ssbCol)) c2n(dat[[ssbCol]]) else NA_real_,
    F              =if (!is.na(fCol)) c2n(dat[[fCol]]) else NA_real_,
    recruits       =if (!is.na(recruitsCol)) c2n(dat[[recruitsCol]]) else NA_real_,
    catches        =if (!is.na(catchesCol)) c2n(dat[[catchesCol]]) else NA_real_,
    landings       =if (!is.na(landingsCol)) c2n(dat[[landingsCol]]) else NA_real_,
    discards       =if (!is.na(discardsCol)) c2n(dat[[discardsCol]]) else NA_real_,
    stringsAsFactors= FALSE)

  out$catch=with(out, ifelse(
    is.finite(catches), catches,
    landings + ifelse(is.finite(discards), discards, 0)))

  out=out[is.finite(out$year),
          c("sid", "assYr", "year", "SSB", "F", "recruits", "catch")]
  out[order(out$year), ]}

sagFn<-function(sid, assYr, data="summary", purpose="Advice") {
  tryCatch(
    withRetries(
      suppressMessages(getSAG(stock=sid, year=assYr, data=data, purpose=purpose)),
      label=paste0("getSAG/", data, " (", sid, ", ", assYr, ")"),
      n=max(2L, apiRetries)),
    error=function(e) {
      warning(conditionMessage(e), call.=FALSE)
      NULL})}

refptCols=c(
  "FLim", "Fpa", "Bpa", "Blim", "FMSY", "BMSY", "MSYBtrigger", "MSYBescapement",
  "Fmanagement", "Bmanagement", "TerminalYear", "FAge", "ModelType", "ModelName")

tidyRefpts<-function(dat, sid, assYr) {
  if (is.null(dat) || !is.data.frame(dat) || nrow(dat) == 0) return(NULL)
  r=dat[nrow(dat), , drop=FALSE]
  out=data.frame(
    sid            =sid,
    assYr=assYr,
    stringsAsFactors=FALSE)
  for (cn in refptCols) {
    out[[cn]]=if (cn %in% names(r)) c2n(r[[cn]]) else NA_real_}
  if ("StockKeyLabel" %in% names(r)) out$stock_key_label=as.character(r$StockKeyLabel)
  out}

adviceYear<-function(x) {
  x=as.character(x)
  d=suppressWarnings(as.Date(sub("T.*", "", x)))
  if (all(is.na(d)) && any(grepl("/", x))) {
    d=suppressWarnings(as.Date(x, format="%d/%m/%Y"))}
  as.integer(format(d, "%Y"))}

tidyAdvice<-function(adv, assYr) {
  if (is.null(adv) || !is.data.frame(adv) || nrow(adv) == 0) return(NULL)
  sidCol=getCol(adv, c("stockcode", "stockkeylabel"))
  if (is.na(sidCol)) return(NULL)

  out=data.frame(
    sid            =as.character(adv[[sidCol]]),
    assYr=assYr,
    advice_key     =if ("adviceKey" %in% names(adv)) adv$adviceKey else NA_integer_,
    advice_year    =adviceYear(adv$adviceApplicableFrom),
    advice_until   =adviceYear(adv$adviceApplicableUntil),
    advice         =if ("adviceValue" %in% names(adv)) c2n(adv$adviceValue) else NA_real_,
    advice_unit    =if ("adviceValueUnit" %in% names(adv)) as.character(adv$adviceValueUnit) else NA_character_,
    wanted_catch   =if ("wantedCatch" %in% names(adv)) c2n(adv$wantedCatch) else NA_real_,
    advice_type    =if ("adviceType" %in% names(adv)) as.character(adv$adviceType) else NA_character_,
    advice_category=if ("adviceCategory" %in% names(adv)) as.character(adv$adviceCategory) else NA_character_,
    advice_sentence=if ("adviceSentence" %in% names(adv)) as.character(adv$adviceSentence) else NA_character_,
    advice_doi     =if ("adviceDOI" %in% names(adv)) as.character(adv$adviceDOI) else NA_character_,
    stringsAsFactors=FALSE)

  out=out[!is.na(out$sid) & nzchar(out$sid), ]
  out}

tidyCatch<-function(dat, sid, assYr, advice_key) {
  if (is.null(dat) || !is.data.frame(dat) || nrow(dat) == 0) return(NULL)
  lbl=getCol(dat, c("ak_label"))
  valCol=getCol(dat, c("value"))
  yrCol=getCol(dat, c("yearlabel"))
  purpCol=getCol(dat, c("cs_purpose", "cS_Purpose"))
  if (is.na(valCol)) return(NULL)

  out=data.frame(
    sid            =sid,
    assYr=assYr,
    advice_key     =advice_key,
    scenario_purpose=if (!is.na(purpCol)) as.character(dat[[purpCol]]) else NA_character_,
    scenario_basis  =if ("cS_Basis" %in% names(dat)) as.character(dat$cS_Basis) else NA_character_,
    variable       =if (!is.na(lbl)) as.character(dat[[lbl]]) else NA_character_,
    advice_year    =if (!is.na(yrCol)) c2n(dat[[yrCol]]) else NA_real_,
    value          =c2n(dat[[valCol]]),
    unit           =if ("unit" %in% names(dat)) as.character(dat$unit) else NA_character_,
    stringsAsFactors=FALSE)

  out=out[!is.na(out$variable), ]
  out}

sagKable<-function(x, caption, align=NULL) {
  if (is.null(align)) align=rep("l", ncol(x))
  tab=knitr::kable(x, caption=caption, align=align, format="html", row.names=FALSE)
  if (requireNamespace("kableExtra", quietly=TRUE)) {
    kableExtra::kable_styling(
      tab,
      bootstrap_options=c("striped", "hover", "condensed"),
      full_width=FALSE,
      position="left")
  } else tab
}

sagSummary<-function() {
  nGrid=nrow(grid)
  nStocks=length(unique(grid$sid))
  nTsPairs=if (exists("sag_ts") && nrow(sag_ts) > 0) {
    nrow(unique(sag_ts[, c("sid", "assYr"), drop=FALSE]))
  } else 0L
  nTsRows=if (exists("sag_ts")) nrow(sag_ts) else 0L
  nRefpts=if (exists("sag_refpts")) nrow(sag_refpts) else 0L
  nAdvice=if (exists("sag_advice")) nrow(sag_advice) else 0L
  nAdviceVal=if (exists("sag_advice")) sum(is.finite(sag_advice$advice)) else 0L
  nScenarios=if (exists("sag_scenarios") && nrow(sag_scenarios) > 0) {
    nrow(sag_scenarios)
  } else NA_integer_

  pct=function(n, d) if (d > 0) sprintf("%.1f%%", 100 * n / d) else "—"

  out=data.frame(
    Metric=c(
      "Assessment years",
      "Stock–assessment pairs",
      "Unique stocks",
      "Time series (stock–assessments)",
      "Time series rows",
      "Reference points",
      "Advice records",
      "Advice values (non-missing)",
      "Catch scenarios"),
    Count=c(
      length(assessYrs),
      nGrid,
      nStocks,
      nTsPairs,
      nTsRows,
      nRefpts,
      nAdvice,
      nAdviceVal,
      if (is.na(nScenarios)) NA_integer_ else nScenarios),
    Coverage=c(
      paste(min(assessYrs), "–", max(assessYrs), sep=""),
      pct(nTsPairs, nGrid),
      "—",
      paste0(format(nTsPairs, big.mark=","), " / ", format(nGrid, big.mark=",")),
      "—",
      pct(nRefpts, nGrid),
      pct(nAdvice, nGrid),
      pct(nAdviceVal, nAdvice),
      if (is.na(nScenarios)) "not fetched" else "fetched"),
    stringsAsFactors=FALSE)
  out$Count=ifelse(is.na(out$Count), "—", format(out$Count, big.mark=",", scientific=FALSE))
  out
}

sagObjects<-function() {
  data.frame(
    Object=c(
      "sag_ts", "sag_refpts", "sag_advice", "sag_stock", "sag_scenarios",
      "stockIndex", "grid", "rawList", "yrFrom", "yrTo"),
    Description=c(
      "Long time series: SSB, F, recruits, catch by calendar year",
      "Reference points (Blim, FMSY, MSYBtrigger, …) per stock × assessment year",
      "Catch advice and metadata from Advice View",
      "One row per stock × assessment year: refpts + primary advice",
      "Catch scenario tables (empty if scenarios not fetched)",
      "Stock list metadata from getListStocks()",
      "Stock × assessment year query grid",
      "Raw SAG summary tables (all columns)",
      "First assessment year in extract",
      "Last assessment year in extract"),
    stringsAsFactors=FALSE)
}
```

```{r stock-list}
loadStockListFromRdata<-function() {
  if (!reuseStockList || !file.exists(rdataFile)) return(NULL)
  env=new.env()
  load(rdataFile, envir=env)
  if (!all(c("grid", "stockIndex") %in% ls(env))) return(NULL)
  stockIndex=get("stockIndex", env)
  list(
    grid=get("grid", env),
    stockIndex=stockIndex,
    stockLists=split(stockIndex, stockIndex$assYr))
}

fetchStockList<-function() {
  stockLists=lapply(assessYrs, function(yr) {
    sl=withRetries(
      suppressMessages(getListStocks(year=yr)),
      label=paste0("getListStocks(", yr, ")"))
    if (!is.null(statusFilter) && nzchar(statusFilter) && "Status" %in% names(sl)) {
      sl=subset(sl, Status == statusFilter)
    }
    sl$assYr=yr
    sl})
  names(stockLists)=assessYrs

  stockIndex=do.call(rbind, lapply(stockLists, function(x) {
    cols=c("StockKeyLabel", "StockDescription", "SpeciesName", "AssessmentYear",
           "assYr", "AssessmentKey", "Status", "LinkToAdvice")
    x[, cols[cols %in% names(x)], drop=FALSE]}))
  names(stockIndex)[1]="sid"

  grid=unique(stockIndex[, c("sid", "assYr")])
  grid=grid[order(grid$assYr, grid$sid), ]
  list(grid=grid, stockIndex=stockIndex, stockLists=stockLists)
}

stockData=tryCatch(
  fetchStockList(),
  error=function(e) {
    cached=loadStockListFromRdata()
    if (!is.null(cached)) {
      warning(
        "Could not reach ICES SAG API; reusing stock list from:\n  ", rdataFile,
        "\nOriginal error: ", conditionMessage(e),
        call.=FALSE)
      cached
    } else {
      stop(
        "Could not reach ICES SAG API and no saved stock list to fall back on.\n",
        "  Expected: ", rdataFile, "\n",
        "Check network/VPN, increase timeouts (connect_timeout_s, request_timeout_s), ",
        "or run again when sag.ices.dk is reachable.\n",
        "Original error: ", conditionMessage(e),
        call.=FALSE)
    }
  })

grid=stockData$grid
stockIndex=stockData$stockIndex
stockLists=stockData$stockLists
```

```{r extract}
# Loop over stock × assessment year (avoids mdply empty bind when all calls fail).
sagParts=vector("list", nrow(grid))
rawParts=vector("list", nrow(grid))

for (i in seq_len(nrow(grid))) {
  sid=grid$sid[i]
  ay =grid$assYr[i]
  raw=sagFn(sid, ay)
  if (is.null(raw) || !is.data.frame(raw) || nrow(raw) == 0) next

  if ("Year" %in% names(raw)) names(raw)[names(raw) == "Year"]="year"
  rawParts[[i]]=data.frame(sid=sid, assYr=ay, raw, check.names=FALSE)
  sagParts[[i]]=tidySmry(raw, sid, ay)}

rawParts=rawParts[!vapply(rawParts, is.null, logical(1))]
sagParts=sagParts[!vapply(sagParts, is.null, logical(1))]

rawList=if (length(rawParts)) do.call(rbind, rawParts) else data.frame()

if (length(sagParts) == 0) {
  sag_ts=data.frame(
    sid=character(), assYr=integer(), year=integer(),
    SSB=double(), F=double(), recruits=double(), catch=double(),
    stringsAsFactors=FALSE
  )
  warning("No SAG data retrieved — check network/API access and stock list.")
} else {
  sag_ts=do.call(rbind, sagParts)
  sag_ts=sag_ts[order(sag_ts$sid, sag_ts$assYr, sag_ts$year), ]
  rownames(sag_ts)=NULL
}
```

```{r extract-refpts}
refptParts=vector("list", nrow(grid))

for (i in seq_len(nrow(grid))) {
  sid=grid$sid[i]
  ay =grid$assYr[i]
  raw=sagFn(sid, ay, data="refpts")
  refptParts[[i]]=tidyRefpts(raw, sid, ay)}

refptParts=refptParts[!vapply(refptParts, is.null, logical(1))]

if (length(refptParts) == 0) {
  sag_refpts=data.frame(sid=character(), assYr=integer())
} else {
  sag_refpts=do.call(rbind, refptParts)
  rownames(sag_refpts)=NULL
}

# Link to advice sheet from stock index where available
if ("LinkToAdvice" %in% names(stockIndex)) {
  links=unique(stockIndex[, c("sid", "assYr", "LinkToAdvice")])
  sag_refpts=merge(sag_refpts, links, by=c("sid", "assYr"), all.x=TRUE)
}
```

```{r extract-advice}
adviceParts=lapply(assessYrs, function(ay) {
  adv=tryCatch(
    withRetries(
      suppressMessages(getAdviceViewRecord(year=ay)),
      label=paste0("getAdviceViewRecord(", ay, ")")),
    error=function(e) {
      warning("getAdviceViewRecord failed (", ay, "): ", conditionMessage(e),
              call.=FALSE)
      NULL})
  if (is.null(adv)) return(NULL)
  adv=tidyAdvice(adv, ay)
  if (is.null(adv)) return(NULL)
  adv[adv$sid %in% grid$sid[grid$assYr == ay], , drop=FALSE]})

adviceParts=adviceParts[!vapply(adviceParts, is.null, logical(1))]

if (length(adviceParts) == 0) {
  sag_advice=data.frame(
    sid=character(), assYr=integer(), advice_year=integer(),
    advice=double(), stringsAsFactors=FALSE)
} else {
  sag_advice=do.call(rbind, adviceParts)
  rownames(sag_advice)=NULL
  sag_advice=sag_advice[order(sag_advice$sid, sag_advice$assYr, sag_advice$advice_year), ]
}
```

```{r extract-advice-scenarios, eval=fetch_scenarios}
scenarioParts=vector("list", nrow(sag_advice))

for (i in seq_len(nrow(sag_advice))) {
  key=sag_advice$advice_key[i]
  if (!is.finite(key)) next
  sid=sag_advice$sid[i]
  ay =sag_advice$assYr[i]
  cs=tryCatch(
    withRetries(
      suppressMessages(getCatchScenariosTable(adviceKey=key)),
      label=paste0("getCatchScenariosTable(", sid, ", key ", key, ")"),
      n=max(2L, apiRetries)),
    error=function(e) {
      warning("getCatchScenariosTable failed (", sid, ", key ", key, "): ",
              conditionMessage(e), call.=FALSE)
      NULL})
  scenarioParts[[i]]=tidyCatch(cs, sid, ay, key)}

scenarioParts=scenarioParts[!vapply(scenarioParts, is.null, logical(1))]

if (length(scenarioParts) == 0) {
  sag_scenarios=data.frame()
} else {
  sag_scenarios=do.call(rbind, scenarioParts)
  rownames(sag_scenarios)=NULL
}
```

## Results

```{r summary, include=TRUE, echo=FALSE, results="asis"}
cat(sprintf(
  "**Output file:** `%s`\n\n",
  normalizePath(rdataFile, winslash="/", mustWork=FALSE)))
print(sagKable(
  sagSummary(),
  caption=sprintf("Extraction summary (%d–%d)", yrFrom, yrTo),
  align=c("l", "r", "r")))
cat("\n\n")
print(sagKable(
  sagObjects(),
  caption="Objects saved in the RData file",
  align=c("l", "l", "l")))
```

```{r save}
if (!exists("sag_refpts")) {
  sag_refpts=data.frame(sid=character(), assYr=integer())
}
if (!exists("sag_advice")) {
  sag_advice=data.frame(
    sid=character(), assYr=integer(), advice_year=integer(),
    advice=double(), stringsAsFactors=FALSE)
}

# Combined stock × assessment table (refpts + primary catch advice)
advPrimary=sag_advice
if (exists("sag_advice") && nrow(sag_advice) > 0 &&
    any(duplicated(paste(sag_advice$sid, sag_advice$assYr)))) {
  advPrimary=do.call(rbind, lapply(
    split(sag_advice, paste(sag_advice$sid, sag_advice$assYr)),
    function(x) {
      x=x[order(x$advice_year), , drop=FALSE]
      x[1, , drop=FALSE]
    }))}

advCols=c("sid", "assYr", "advice_year", "advice", "advice_unit",
          "wanted_catch", "advice_type", "advice_sentence", "advice_key")
advCols=advCols[advCols %in% names(advPrimary)]

sag_stock=merge(unique(grid), sag_refpts, by=c("sid", "assYr"), all.x=TRUE)
if (exists("advPrimary") && nrow(advPrimary) > 0 && length(advCols) > 2) {
  sag_stock=merge(sag_stock, advPrimary[, advCols, drop=FALSE],
                  by=c("sid", "assYr"), all.x=TRUE)
}

if (!exists("sag_scenarios")) {
  sag_scenarios=data.frame()}

save(
  sag_ts, sag_refpts, sag_advice, sag_stock, sag_scenarios,
  stockIndex, grid, rawList, yrFrom, yrTo,
  file=rdataFile)
```

## Coverage by assessment year

```{r preview, include=TRUE, echo=FALSE, results="asis"}
coverage=ddply(grid, .(assYr), function(g) {
  got=unique(sag_ts$sid[sag_ts$assYr == g$assYr[1]])
  nReq=nrow(g)
  nGot=length(intersect(g$sid, got))
  data.frame(
    assYr=g$assYr[1],
    stocks_requested=nReq,
    stocks_with_series=nGot,
    missing=nReq - nGot,
  pct_with_series=round(100 * nGot / nReq, 1),
  stringsAsFactors=FALSE)
})
coverage=coverage[order(coverage$assYr, decreasing=TRUE), ]
print(sagKable(
  coverage,
  caption="Stocks with at least one time-series row in SAG summary",
  align=c("r", "r", "r", "r", "r")))
```

## Example time series

Northeast Atlantic mackerel (`mac.27.nea`): each coloured line is the series as published in that assessment year.

```{r example-plot, fig.height=7, fig.width=9, include=TRUE, eval=requireNamespace("ggplot2", quietly=TRUE)}
if (requireNamespace("ggplot2", quietly=TRUE) && nrow(sag_ts) > 0) {
  exSid=intersect("mac.27.nea", unique(sag_ts$sid))[1]
  if (!is.na(exSid)) {
    ex=subset(sag_ts, sid == exSid)
    exLong=tidyr::pivot_longer(
      ex,
      cols=c("SSB", "F", "recruits", "catch"),
      names_to="variable",
      values_to="value")
    varLabs=c(SSB="SSB", F="F", recruits="Recruits", catch="Catch")
    exLong$variable=factor(exLong$variable, levels=names(varLabs), labels=varLabs)

    ggplot2::ggplot(exLong, ggplot2::aes(year, value, colour=factor(assYr))) +
      ggplot2::geom_line(linewidth=0.7, alpha=0.85) +
      ggplot2::facet_wrap(~variable, scales="free_y", ncol=1) +
      ggplot2::scale_colour_viridis_d(option="C", name="Assessment\nyear") +
      ggplot2::labs(
        title="Northeast Atlantic mackerel",
        subtitle="SAG summary by assessment year",
        x="Calendar year",
        y=NULL) +
      ggplot2::theme_minimal(base_size=11) +
      ggplot2::theme(
        legend.position="right",
        panel.grid.minor=ggplot2::element_blank(),
        strip.text=ggplot2::element_text(face="bold"))
  }
}
```
